---
title: 'Grassmann analysis'
author: "Toka Othman"
output:
  word_document: default
  html_document: default
  pdf_document: default
---
Re-analysis of Grassmann et al (2025)
Borrelia transcriptome comparison between WT vs KO and WT vs R39A mutants


# Part 1. load libraries; set working directory


```{r}
library(tidyverse)
library(readxl)
library(dplyr)
library(EnhancedVolcano)
library(ggrepel)
library(shiny)
library(bslib)
library(modeldata)
library(DataExplorer)
library(plotly)
library(ggplot2)
setwd("~/QiuLab-work/")
```

# Part 2. Read data table, reshape into long tables

```{r}
x <- read_excel("mmi70036-sup-0002-tables2.xlsx", sheet = 3)
colnames(x)

x.long <- x |>
    select(1, 8:14) |>
    pivot_longer(2:8, names_to = "treat", values_to = "rfpkm")  

x.long <- x.long |>
    mutate(rep = as.numeric(str_remove(treat, "^.*_")))

x.long <- x.long |>
    mutate(treat2 = str_remove(treat, "_\\d$"))

```


# Part 3. Plot histogram to show distributions

```{r}
x.long |>
    ggplot(aes(rfpkm+1, fill = as.factor(rep))) +
    geom_histogram(bins = 100) +
    scale_x_log10() +
    facet_wrap(~treat2, nrow=2)
```


# Part 4. plot volcano plot

```{r}
library(EnhancedVolcano)
library(ggrepel)

EnhancedVolcano(
    toptable = x,
    lab = x$GeneID,
    x = "log2FoldChange",
    y = "pvalue",
    title = "WT_in_vitro vs ΔbosR _in_vitro",
    pCutoff = 0.05,
    FCcutoff = 1,
    pointSize = 2.0,
    labSize = 3.0)
```
# Part 5. read data frame

```{r}
y <- read_excel("mmi70036-sup-0003-tables3.xlsx", sheet = 3)

```

# Part 6. select desired data

```{r}

in_vitro <- x[, c(1, 8, 9, 10)]
colnames(in_vitro) <- c("GeneID", "WT_in_vitro_1", "WT_in_vitro_2", "WT_in_vitro_3")


DMC <- y[, c(1, 8:13)]
colnames(DMC) <- c("GeneID","WT_DMC1","WT_DMC2","WT_DMC3","WT_DMC4","WT_DMC5","WT_DMC6")


WT <- merge(in_vitro, DMC , by = "GeneID")

```

# Part 7. calculate fold change 
read counts need to be in log2+1; and then take the mean for WT & 

```{r}

FC <- WT %>%
mutate(in_vitro_mean = rowMeans(select(., 2:4)),
 DMC_mean = rowMeans(select(., 5:10)),
 fold_change = DMC_mean / in_vitro_mean,
 log2FC = log2(fold_change))

FC |>
  ggplot(aes(x = log2FC)) +
  geom_histogram(bins = 100) +
  geom_vline(xintercept = 0, col = "red") +
  theme_bw()

```

# Part 8. Calculate pvalue

```{r}

FC <- FC %>%
  rowwise() %>%
  mutate(pvalue = t.test(c_across(2:4), c_across(5:10))$p.value) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(pvalue, method = "BH"))

```

```{r}
FC_clean <- FC %>% filter(is.finite(log2FC),is.finite(pvalue))

# show BB_Q67, methyl-transferase
gene.sel <- FC_clean |> filter(GeneID == 'BB_Q67') 
#|> select(in_vitro_mean, DMC_mean, fold_change, pvalue, p_adj)
```

# Part 9. plot volcano plot

```{r}

EnhancedVolcano(
  toptable = FC_clean,
  lab = FC_clean$GeneID,
  x = "log2FC",
  y = "pvalue",
  title = "WT_in_vitro vs WT_in_DMC",
  pCutoff = 0.05,
  FCcutoff = 1,
  pointSize = 2.0,
  labSize = 3.0) 

FC_clean |>
  ggplot(aes(log2FC, p_adj)) +
  geom_point(shape = 1) +
  geom_point(data = gene.sel, col = "red") +
  scale_y_log10() +
  geom_vline(xintercept = 0, linetype = 2) +
  geom_hline(yintercept = 1e-3, col = "red") +
  theme_bw()
#  geom_point(data = gene.sel, size = 3, col = "orange" )

```